Viscous Transport in Eroding Porous Media

### 2019-10-03 docs/transport/erosion_transport.tex
NM
working on Nick's revisions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Single Body Close to a Wall}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
First, consider a single eroding body close to a solid wall with the
shear flow
\begin{align}
  \UU(\xx)=\left[
  \begin{array}{c}
    y+1 \\ 0
  \end{array}
  \right],
\end{align}
imposed on the outer wall. Mitchell and Spagnolie~\cite{mit-spa2017}
performed a similar three-dimensional experiment using a second-order
quadrature method.  They initialized the \nick{\st{eroding} body \st{so that}
with its center located} $1.5$ radii above the solid wall.  In two
dimensions, the trapezoid rule is spectrally accurate, and an analogous
initial configuration is possible. However, unless the number of
discretization points is increased, the trapezoid rule becomes obsolete
if the eroding body is initialized closer to the solid wall. Therefore,
to simulate erosion near a solid wall at reasonable levels of
discretization, we use our hybrid method that combines the trapezoid
rule and the Barycentric quadrature rule.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Medium Porosity}
\label{sec:Eroding20}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider 20 eroding bodies that are discretized with $N_\iin=256$
points.  The background flow is the Hagen-Poiseuille flow
\begin{align}
  \UU(\xx)=U \left[
  \begin{array}{c}
    1-y^2 \\ 0
  \end{array}
  \right],
\end{align}
where the flow rate $U$ is chosen so that the pressure drop from $x=-2$
to $x=2$ is held fixed at 8. Therefore, $U=1$ once all the bodies have
vanished.  Snapshots of the bodies at four equispaced times are shown in
Figure~\ref{fig:Eroding20vort}.  The color is the vorticity, \nick{which is proportional to the erosion rate on solid boundaries}.
Initially, several of the eroding bodies are closer to the outer wall than the $5h$ threshold \nick{required \st{to guarantee that} for} the trapezoid rule to achieve machine
precision.  In particular, the distance between bodies 1, 6, 13, and 15
and the outer wall is $1.3h$, $2.9h$, $2.8h$, and $1.3h$, respectively,
where $h$ is the arclength spacing of the outer wall $\Gamma$.  In
addition, the distance between several pairs of eroding bodies,
including 1 \& 6, 3 \& 9, 6 \& 8, and 14 \& 18, is too small to be
accurately resolved with the trapezoid rule.  By using the Barycentric
quadrature rule, the interaction between these nearly-touching bodies
can be accurately resolved \nick{\st{at the prescribed resolution} to the desired accuracy}, and erosion can
be simulated until all the bodies have vanished.

@@

\nick{\st{Similar to our previous work,}} Erosion causes the space between nearly
touching bodies to quickly expand, and flat faces develop along the
regions of near contact.  This qualitative behavior is seen in Figure~\ref{fig:Eroding20vort} between bodies 3 \& 4, 15 \& 16, and
\nick{was also witnessed in previous work}.  
However, by resolving the interaction between bodies that are
much closer together than in previous work, we observe that very little erosion occurs between certain pairs of bodies, at least initially.  
For instance the opening between bodies 1 \& 6, 3 \& 9, and 5
\& 13 grow much slower than the opening between bodies 15 \& 16.  A
common feature of the pores that grow slowly is that they are nearly
perpendicular to the main flow direction, \nick{resulting} in a small
erosion rate.

Following our previous work~\cite{qua-moo2018}, we begin by analyzing
the effect of erosion on the area fraction and the flow rate.  In
Figure~\ref{fig:Eroding20flowrate}(a), we plot the area fraction as a
function of normalized time.  The general trend of the area fraction
resembles our previous work~\cite{qua-moo2018} (Figure 10(a)), but with
a larger initial area fraction.  In
Figure~\ref{fig:Eroding20flowrate}(b), we plot the flow rate $U$
required to maintain a constant pressure drop across the channel.
Again, the trend of $U$ resembles that of our previous
work~\cite{qua-moo2018} (Figure 10(b)), except that the initial flow
rate is an order of magnitude smaller because of the larger initial area
fraction.  Starting around normalized time $0.2$,
Figure~\ref{fig:Eroding20flowrate}(b) \nick{is roughly linear which indicates
that the flow rate can be written as a power law.  Using a line of best
fit, the flow rate is approximately 
\begin{align} 
  U \approx \left(\frac{t}{t_f}\right)^{3.43},
\end{align}
which is the dashed line in} Figure~\ref{fig:Eroding20flowrate}(b).

\todo[inline]{WAIT! Fig 7b is a semi-log plot with the x-axis linear and the y-axis logarithmic. Therefore, a straight line indicates exponential increase, not a power law.}

@@

In Figure~\ref{fig:Eroding20vort}, we observe that erosion creates a
network of channels from the inlet to outlet where the velocity and
vorticity, and therefore erosion rate, are much larger than other
regions.  These channels can be further visualized by considering the
streamlines.  In Figure~\ref{fig:Eroding20tracer}, we freeze the
geometry at the second snapshot from Figure~\ref{fig:Eroding20vort} and plot 200 streamlines that are initialized at $x=-1$ and equally
spaced in $y \in (-1,1)$.  The streamlines are shown at five different
times, and the final plot is a zoom in of the lower right quadrant of
the fifth time step.  Since we use a high-order quadrature rule and time
stepping method, we resolve streamlines that come very close to the
eroded bodies.  There are three clear regions where the streamlines are \nick{most}
concentrated, \nick{\st{and this}} corresponding to the regions of highest velocity.  Two of these regions are located between the bodies and the solid walls at
$y=\pm 1$, and the third cuts through the porous region with the upper
part of the channel formed by bodies 1, 4, 6, and 8.  
\todo[inline]{Right now, this description has the reader looking at Fig. 8, but the bodies are not numbered in that figure, so they don't know which ones are 1, 4, 6, 8 without flipping back and forth with the previous figures to try to match corresponding bodes.}
Since the flow is fastest in these regions, the shearing is largest, and this causes the
channels to continue to open fastest \nick{\st{which can be} as} observed in
Figure~\ref{fig:Eroding20vort}.

@@

Next, we use the streamlines to compute the tortuosity and the anomalous
dispersion rate\nick{s} of eroding geometries.  To achieve statistical
convergence, we use $N_p = 1000$ streamlines~\cite{bel-sal-rin1992}. To
compute the tortuosity, we require the velocity at the inlet $x=-1$.
These normalized velocities are plotted in
Figure~\ref{fig:Eroding20tort}(a) for the eroded geometry at porosity
$\phi = 62.9\%$ (Figure~\ref{fig:Eroding20tort}(c)).  The velocities are
similar to the work of Matyka et al.~\cite{mat-kha-koz2008} (see Figure
4(a)), except that our cross-section, by construction, does not cut
through any of the grains.  Next, in Figure~\ref{fig:Eroding20tort}(b),
we plot the local tortuosity~\eqref{eqn:localTort} by calculating the
relative length of each streamline as it traverses the channel from
$x=-1$ to $x=1$.  The local tortuosity ranges from 1 to 1.27, meaning
that one of the streamlines is 27\% longer than it would have been if
the grains were absent.  The average streamline is 9.79\% longer
\nick{\st{than if the grains had been absent}, \st{so the} or equivalently the tortuosity of the \st{eroded channel} geometry is $1.098$}.  Again, comparing the local tortuosity to
Matyka~\cite{mat-kha-koz2008} (Figure 4(b)), the results are
qualitatively similar. However, since our initial cross-section does not
cut through the grains, the local tortuosity does not have any gaps.
\nick{Discontinuities in local tortuosity \st{happens} occur} when nearby
streamlines diverge \nick{\st{on opposite sides of a grain} to circumvent a grain}.  In
Figure~\ref{fig:Eroding20tort}(c), we plot pairs of streamlines \nick{associate with the ten largest jumps in the local tortuosity, with each pair of corresponding streamlines plotted in the same color.}

@@

In Figure~\ref{fig:Eroding20Transport}(a), we plot the tortuosity as a
function of the porosity. The initial porosity is $\phi = 37.68\%$, and
the initial tortuosity is $T = 1.16$.  The tortuosity is computed with both
the length of the streamlines~\eqref{eqn:tortuosity1} (red stars) and
using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in Figure~\ref{fig:Eroding20tort}(c).
The two tortuosity formulas give similar results, and any discrepancy
can be accounted for by slow regions of recirculation and from applying
quadrature to compute the tortuosity.  As the bodies erode, wide
channels form where streamlines undergo only minor vertical deflections,
and this explains why the tortuosity eventually decreases with porosity.
We computed lines of best fit using the porosity-tortuosity
models~\eqref{eqn:tortuosityModels} and \nick{found the power law to give} the
smallest error.  The  black dashed line in
Figure~\ref{fig:Eroding20Transport}(a) is the line of best fit
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.2064}
\end{align}
with a root-mean-square error of $5.90 \times 10^{-3}$.  Interestingly,
at the low porosities, the tortuosity initially increases. This increase
occurs because in the absence of erosion (left plot in
Figure~\ref{fig:Eroding20vort}), many of the streamlines, such as those
initialized between bodies 15 \& 18, only perform minor deflections to
pass through the narrow regions, albeit, very slowly.  However, as
erosion starts to open the channels, the streamlines deflect into the
fast regions, such as the region above body 11, and this increases the
amount of vertical deflection, and therefore the tortuosity.

@@

In Figure~\ref{fig:Eroding20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  The spreading is
computed \nick{\st{at} for} seven geometries of different porosities that are formed
during the erosion process.  So that the spreading reaches a statistical
equilibrium, we use the reinsertion algorithm described in
Section~\ref{sec:dispersion} to form sufficiently long trajectories.
For all the reported porosities, the particle dispersion exhibits two \nick{distinct
power \st{low} law} regimes.  Initially, the dispersion is ballistic
($\sigma_\lambda \sim t$) since individual fluid particles have not yet
explored enough space to significantly alter their velocity.  However,
once the particles have been subjected to a range of velocities, their
dispersion slows \nick{\st{down}}, and we observe super-dispersive ($\sigma_\lambda
\sim t^\alpha$, $\alpha \in (1/2,1)$) behavior over at least one order
of magnitude in time.  Before any erosion takes place, the anomalous
dispersion coefficient is $\alpha = 0.56$.  Then, as the grains begin to
erode, the dispersion rate grows towards the ballistic regime that
occurs in the absence of grains.  The monotonic increase in dispersion
with respect to the porosity is explained by the onset of channels where
many tracers experience less variability in their velocities.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Single Body Close to a Wall}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
First, consider a single eroding body close to a solid wall with the
shear flow
\begin{align}
  \UU(\xx)=\left[
  \begin{array}{c}
    y+1 \\ 0
  \end{array}
  \right],
\end{align}
imposed on the outer wall. Mitchell and Spagnolie~\cite{mit-spa2017}
performed a similar three-dimensional experiment using a second-order
quadrature method.  They initialized the body with its center located
$1.5$ radii above the solid wall.  In two dimensions, the trapezoid rule
is spectrally accurate, and an analogous initial configuration is
possible. However, unless the number of discretization points is
increased, the trapezoid rule becomes obsolete if the eroding body is
initialized closer to the solid wall. Therefore, to simulate erosion
near a solid wall at reasonable levels of discretization, we use our
hybrid method that combines the trapezoid rule and the Barycentric
quadrature rule.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Medium Porosity}
\label{sec:Eroding20}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider 20 eroding bodies that are discretized with $N_\iin=256$
points.  The background flow is the Hagen-Poiseuille flow
\begin{align}
  \UU(\xx)=U \left[
  \begin{array}{c}
    1-y^2 \\ 0
  \end{array}
  \right],
\end{align}
where the flow rate $U$ is chosen so that the pressure drop from $x=-2$
to $x=2$ is held fixed at 8. Therefore, $U=1$ once all the bodies have
vanished.  Snapshots of the bodies at four equispaced times are shown in
Figure~\ref{fig:Eroding20vort}.  The color is the vorticity, whose
absolute value is proportional to the erosion rate on solid boundaries.
Initially, several of the eroding bodies are closer to the outer wall
than the $5h$ threshold required for the trapezoid rule to achieve
machine precision.  In particular, the distance between bodies 1, 6, 13,
and 15 and the outer wall is $1.3h$, $2.9h$, $2.8h$, and $1.3h$,
respectively, where $h$ is the arclength spacing of the outer wall
$\Gamma$.  In addition, the distance between several pairs of eroding
bodies, including 1 \& 6, 3 \& 9, 6 \& 8, and 14 \& 18, is too small to
be accurately resolved with the trapezoid rule.  By using the
Barycentric quadrature rule, the interaction between these
nearly-touching bodies can be accurately resolved to the desired
accuracy, and erosion can be simulated until all the bodies have
vanished.

@@

Erosion causes the space between nearly touching bodies to quickly
expand, and flat faces develop along the regions of near contact.  This
qualitative behavior is seen in Figure~\ref{fig:Eroding20vort} between
bodies 3 \& 4, 15 \& 16, and was also observed in previous work.
However, by resolving the interaction between bodies that are much
closer together than in previous work, we observe that very little
erosion occurs between certain pairs of bodies, at least initially.  For
instance the opening between bodies 1 \& 6, 3 \& 9, and 5 \& 13 grow
much slower than the opening between bodies 15 \& 16.  A common feature
of the pores that grow slowly is that they are nearly perpendicular to
the main flow direction, resulting in a small erosion rate.

Following our previous work~\cite{qua-moo2018}, we begin by analyzing
the effect of erosion on the area fraction and the flow rate.  In
Figure~\ref{fig:Eroding20flowrate}(a), we plot the area fraction as a
function of normalized time.  The general trend of the area fraction
resembles our previous work~\cite{qua-moo2018} (Figure 10(a)), but with
a larger initial area fraction.  In
Figure~\ref{fig:Eroding20flowrate}(b), we plot the flow rate $U$
required to maintain a constant pressure drop across the channel.
Again, the trend of $U$ resembles that of our previous
work~\cite{qua-moo2018} (Figure 10(b)), except that the initial flow
rate is an order of magnitude smaller because of the larger initial area
fraction.  Starting around normalized time $0.2$,
Figure~\ref{fig:Eroding20flowrate}(b) is roughly linear which indicates
that the flow rate can be written as a exponential law.  The line of
best fit is
\begin{align} 
  \log_{10} U \approx 3.43 \frac{t}{t_f} + c, 
\end{align}
$t_f$ is the time required for all the bodies to erode, and $c$ is a
fitting parameter. By construction $U = 1$ when $t=t_f$, so $c=-3.43$,
and the flow rate is approximated as
\begin{align}
  U \approx \exp\left(7.90 \left(\frac{t - t_f}{t_f} \right)\right),
\end{align}
which is the dashed line in Figure~\ref{fig:Eroding20flowrate}(b).

@@

In Figure~\ref{fig:Eroding20vort}, we observe that erosion creates a
network of channels from the inlet to outlet where the velocity and
vorticity, and therefore erosion rate, are much larger than other
regions.  These channels can be further visualized by considering the
streamlines.  In Figure~\ref{fig:Eroding20tracer}, we freeze the
geometry at the second snapshot from Figure~\ref{fig:Eroding20vort} and plot 200 streamlines that are initialized at $x=-1$ and equally
spaced in $y \in (-1,1)$.  The streamlines are shown at five different
times, and the final plot is a zoom in of the lower right quadrant of
the fifth time step.  Since we use a high-order quadrature rule and time
stepping method, we resolve streamlines that come very close to the
eroded bodies.  There are three clear regions where the streamlines are
most concentrated, corresponding to the regions of highest velocity.
Two of these regions are located between the bodies and the solid walls
at $y=\pm 1$, and the third cuts through the porous region with the
upper part of the channel formed by bodies 1, 4, 6, and 8.
\todo[inline]{Right now, this description has the reader looking at Fig.
8, but the bodies are not numbered in that figure, so they don't know
which ones are 1, 4, 6, 8 without flipping back and forth with the
previous figures to try to match corresponding bodes.} Since the flow is
fastest in these regions, the shearing is largest, and this causes the
channels to continue to open fastest as observed in
Figure~\ref{fig:Eroding20vort}.

@@

Next, we use the streamlines to compute the tortuosity and the anomalous
dispersion rates of eroding geometries.  To achieve statistical
convergence, we use $N_p = 1000$ streamlines~\cite{bel-sal-rin1992}. To
compute the tortuosity, we require the velocity at the inlet $x=-1$.
These normalized velocities are plotted in
Figure~\ref{fig:Eroding20tort}(a) for the eroded geometry at porosity
$\phi = 62.9\%$ (Figure~\ref{fig:Eroding20tort}(c)).  The velocities are
similar to the work of Matyka et al.~\cite{mat-kha-koz2008} (see Figure
4(a)), except that our cross-section, by construction, does not cut
through any of the grains.  Next, in Figure~\ref{fig:Eroding20tort}(b),
we plot the local tortuosity~\eqref{eqn:localTort} by calculating the
relative length of each streamline as it traverses the channel from
$x=-1$ to $x=1$.  The local tortuosity ranges from 1 to 1.27, meaning
that one of the streamlines is 27\% longer than it would have been if
the grains were absent.  The average streamline is 9.79\% longer or
equivalently the tortuosity of the geometry is $1.098$.  Again,
comparing the local tortuosity to Matyka~\cite{mat-kha-koz2008} (Figure
4(b)), the results are qualitatively similar. However, since our initial
cross-section does not cut through the grains, the local tortuosity does
not have any gaps.  Discontinuities in local tortuosity occur when
nearby streamlines diverge to circumvent a grain.  In
Figure~\ref{fig:Eroding20tort}(c), we plot pairs of streamlines
associated with the ten largest jumps in the local tortuosity, with each
pair of corresponding streamlines plotted in the same color.

@@

In Figure~\ref{fig:Eroding20Transport}(a), we plot the tortuosity as a
function of the porosity. The initial porosity is $\phi = 37.68\%$, and
the initial tortuosity is $T = 1.16$.  The tortuosity is computed with
both the length of the streamlines~\eqref{eqn:tortuosity1} (red stars)
and using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in Figure~\ref{fig:Eroding20tort}(c).
The two tortuosity formulas give similar results, and any discrepancy
can be accounted for by slow regions of recirculation and from applying
quadrature to compute the tortuosity.  As the bodies erode, wide
channels form where streamlines undergo only minor vertical deflections,
and this explains why the tortuosity eventually decreases with porosity.
We computed lines of best fit using the porosity-tortuosity
models~\eqref{eqn:tortuosityModels} and found the power law to give the
smallest error.  The  black dashed line in
Figure~\ref{fig:Eroding20Transport}(a) is the line of best fit
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.2064}
\end{align}
with a root-mean-square error of $5.90 \times 10^{-3}$.  Interestingly,
at the low porosities, the tortuosity initially increases. This increase
occurs because in the absence of erosion (left plot in
Figure~\ref{fig:Eroding20vort}), many of the streamlines, such as those
initialized between bodies 15 \& 18, only perform minor deflections to
pass through the narrow regions, albeit, very slowly.  However, as
erosion starts to open the channels, the streamlines deflect into the
fast regions, such as the region above body 11, and this increases the
amount of vertical deflection, and therefore the tortuosity.

@@

In Figure~\ref{fig:Eroding20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  The spreading is
computed for seven geometries of different porosities that are formed
during the erosion process.  So that the spreading reaches a statistical
equilibrium, we use the reinsertion algorithm described in
Section~\ref{sec:dispersion} to form sufficiently long trajectories.
For all the reported porosities, the particle dispersion exhibits two
distinct power law regimes.  Initially, the dispersion is ballistic
($\sigma_\lambda \sim t$) since individual fluid particles have not yet
explored enough space to significantly alter their velocity.  However,
once the particles have been subjected to a range of velocities, their
dispersion slows, and we observe super-dispersive ($\sigma_\lambda \sim
t^\alpha$, $\alpha \in (1/2,1)$) behavior over at least one order of
magnitude in time.  Before any erosion takes place, the anomalous
dispersion coefficient is $\alpha = 0.56$.  Then, as the grains begin to
erode, the dispersion rate grows towards the ballistic regime that
occurs in the absence of grains.  The monotonic increase in dispersion
with respect to the porosity is explained by the onset of channels where
many tracers experience less variability in their velocities.

@@

\todo[inline]{Nick made it to here.}
